---
title: "BLR Dissertation Hypothesis 1 and K-Means Analysis and Allometry"
author: "Bonnie Reynolds"
date: '2025'
output:
  pdf_document:
    toc: yes
    number_sections: yes
    highlight: tango
  html_document:
    toc: yes
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load packages

```{r load, out.width = "100%", dpi = 300, echo = TRUE, warning=FALSE}
library(here)
library(StereoMorph)
library(geomorph)
library(wesanderson)
library(tidyverse)
```

## Read data and define sLMs (just leaving this here for safe keeping, don't run it unless making shape outlines)
digitizeImages(image.file='points', 
              shapes.file='shapes', 
              landmarks.ref='landmarks.txt', 
              curves.ref = 'curves.txt'
              )

## Read data and define sLMs
```{r read, out.width = "100%", dpi = 300, echo = TRUE, warning=FALSE}
shapes <- readShapes("shapes")
shapesGM <- readland.shapes(shapes, nCurvePts = c(10, 10, 10, 10))
```

## Read qualitative data
```{r read qual, out.width = "100%", dpi = 300, echo = TRUE, warning=FALSE}
qdata <- read.csv("qdata.csv",
                  header = TRUE,
                  row.names = 1)
```

## Create Kmeans parameters
```{r parameters, echo=TRUE}
# K-Means parameters
max_k <- 10         # Maximum number of clusters to test (Elbow Plot)
nstart_value <- 25  # Number of starts for K-Means
cluster_choices <- c(3, 6, 9)  # Cluster counts to analyze
## Read data and define sLMs
```

## GPA
```{r gp, out.width = "100%", dpi = 300, echo = TRUE, warning=FALSE}
Y.gpa <- gpagen(shapesGM, print.progress = FALSE)
plot(Y.gpa)
```

## Kmeans Cluster Analyses

## K-Means Cluster Analysis (Shape)
```{r kmeans, out.width = "100%", dpi = 300, echo = TRUE, warning=FALSE}
# Prepare GPA coordinates
gpa_flat <- as.data.frame(t(apply(Y.gpa$coords, 3, c)))

# 1. Determine optimal number of clusters (Elbow Method)
wss <- vector()
for (k in 1:max_k) {
  km <- kmeans(gpa_flat, centers = k, nstart = nstart_value)
  wss[k] <- km$tot.withinss
}
plot(1:max_k, wss, type = "b", pch = 19, frame = FALSE,
     xlab = "Number of clusters K",
     ylab = "Total within-clusters sum of squares",
     main = "Elbow Plot for Optimal K")

# 2. K-Means Clustering with specified cluster choices
set.seed(123)
cluster_results <- list()
for (k in cluster_choices) {
  cluster_results[[paste0("k_", k)]] <- kmeans(gpa_flat, centers = k, nstart = nstart_value)
}

# Save assignments to a data frame
assignments_shape <- data.frame(Sample = rownames(gpa_flat))
for (k in cluster_choices) {
  assignments_shape[[paste0("Cluster", k)]] <- cluster_results[[paste0("k_", k)]]$cluster
}

# Show cluster assignments in PDF if you are knitting
knitr::kable(assignments_shape, caption = "K-Means Cluster Assignments (Shape Only)")

# 3. PCA for visualization
pca_result <- prcomp(gpa_flat, scale. = TRUE)
pca_df <- data.frame(pca_result$x[, 1:2])

# Add cluster assignments
for (k in cluster_choices) {
  pca_df[[paste0("Cluster", k)]] <- factor(cluster_results[[paste0("k_", k)]]$cluster)
}

# 4. Plot PCA with pretty cluster colors
library(ggplot2)
for (k in cluster_choices) {
  p <- ggplot(pca_df, aes(x = PC1, y = PC2, color = .data[[paste0("Cluster", k)]])) +
    geom_point(size = 2) +
    theme_minimal() +
    labs(title = paste("K-Means Clustering (", k, " Clusters)", sep = ""), color = "Cluster")
  
  print(p)
}


```

## K-Means Cluster Analysis (Size)

```{r kmeans-size, out.width = "100%", dpi = 300, echo = TRUE, warning=FALSE}
# Reminder to Bonnie - Make sure required libraries are loaded

# Extract Centroid Size
size_data <- data.frame(Size = Y.gpa$Csize)
rownames(size_data) <- dimnames(Y.gpa$coords)[[3]]

# 1. Elbow Method for optimal K
wss <- vector()
for (k in 1:max_k) {
  km <- kmeans(size_data, centers = k, nstart = nstart_value)
  wss[k] <- km$tot.withinss
}
plot(1:max_k, wss, type = "b", pch = 19, frame = FALSE,
     xlab = "Number of clusters K",
     ylab = "Total within-clusters sum of squares",
     main = "Elbow Plot (Size Only)")

# 2. K-Means Clustering on size
set.seed(123)
cluster_results_size <- list()
for (k in cluster_choices) {
  cluster_results_size[[paste0("k_", k)]] <- kmeans(size_data, centers = k, nstart = nstart_value)
}

# Save assignments
assignments_size <- data.frame(Sample = rownames(size_data))
for (k in cluster_choices) {
  assignments_size[[paste0("Cluster", k)]] <- cluster_results_size[[paste0("k_", k)]]$cluster
}

# 3. Visualize with 1D plot (could do more if working with other types of data)(pretty colors again)
for (k in cluster_choices) {
  size_data$Cluster <- factor(cluster_results_size[[paste0("k_", k)]]$cluster)
  p <- ggplot(size_data, aes(x = Cluster, y = Size, color = Cluster)) +
    geom_jitter(width = 0.1, size = 2) +
    theme_minimal() +
    labs(title = paste("K-Means Clustering by Size (K =", k, ")"), x = "Cluster", y = "Centroid Size")
  print(p)
}

# 4. Output table for when knitting
knitr::kable(assignments_size, caption = "K-Means Cluster Assignments (Size Only)")


```

## dataframe model 1
```{r dataframe, out.width = "100%", dpi = 300, echo = TRUE, warning=FALSE}
gdf <- geomorph.data.frame(shape = Y.gpa$coords,
                            size = Y.gpa$Csize,
                            region = qdata$corep) #Level I Ecoregions (per EPA)

# add centroid size to qdata
qdata$csz <- Y.gpa$Csize
```

## Test for Allometry
```{r allometry}
# Basic Allometry Test
allometry_model <- procD.lm(shape ~ size, data = gdf)
summary(allometry_model)

# Plot regression scores
plot(allometry_model, type = "regression", predictor = gdf$size)

```
## Boxplot
```{r box, out.width = "100%", dpi = 300, echo = TRUE, warning=FALSE}
#attributes
csz <- qdata$csz
corep <- qdata$corep

# Custom pink + purple color palette - because women and non-binary people belong in STEM
# Custom pink + purple color palette
pal <- c("#FFB7C5",  # pastel pink
         "#D291BC",  # soft lilac
         "#F7C5E0",  # bubblegum
         "#CBAACB",  # dusty mauve
         "#FFD1DC",  # light pink
         "#E0BBE4",  # soft lavender
         "#FF99C8")  # bold cotton candy
        
#boxplot of Perdiz arrow points by region
csz.reg <- ggplot(qdata, aes(x = corep, y = csz, color = corep)) +
  geom_boxplot() +
  scale_color_manual(values = pal) + 
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1)) +
  theme(legend.position = "none") +
  labs(x = 'Region', y = 'Centroid Size')

# make plot
csz.reg
```

## ANOVA - size then shape
```{r shapesize, out.width = "100%", dpi = 300, echo = TRUE, warning=FALSE}
#size
fit.sz.reg <- procD.lm(size ~ corep,
                       data = gdf,
                       print.progress = FALSE,
                       iter = 9999)

# size
anova(fit.sz.reg)

# pairwise comparison of LS means = which differ? (still size)
sz.reg <- pairwise(fit.sz.reg,
                   groups = qdata$corep)
summary(sz.reg,
        confidence = 0.95,
        test.type = "dist")

# shape
fit.sh.reg <- procD.lm(shape ~ corep,
                       data = gdf,
                       print.progress = FALSE,
                       iter = 9999)

#shape
anova(fit.sh.reg)

# pairwise comparison of LS means = which differ? (now shape)
sh.reg <- pairwise(fit.sh.reg,
                   groups = qdata$corep)
summary(sh.reg,
        confidence = 0.95,
        test.type = "dist")
```

## Mean shapes
```{r mean-shape-comparisons, out.width = "100%", dpi = 300, echo = TRUE, warning=FALSE, message=FALSE}

# Subset landmark coordinates by Core/Periphery group
new.coords <- coords.subset(A = Y.gpa$coords, group = qdata$corep)  # <-- remember to change here
mean_shapes <- lapply(new.coords, mshape)

# Show mean shapes
par(mfrow = c(2, 2))  # remember for later to djust depending on how many groups
for (group in names(mean_shapes)) {
  plot(mean_shapes[[group]], main = paste("Mean Shape:", group))
}
par(mfrow = c(1, 1))

# Show pairwise comparisons one by one
comparison_pairs <- combn(names(mean_shapes), 2, simplify = FALSE)

# Show pairwise comparisons with pink and purple colors (because women and nonbinary people belong in coding), NO numbers, and a adjusted legend
for (pair in comparison_pairs) {
  ref <- mean_shapes[[pair[1]]]
  target <- mean_shapes[[pair[2]]]
  
  # Find plot limits so both shapes fit nicely
  xlim_range <- range(c(ref[,1], target[,1])) * 1.1  # a little padding
  ylim_range <- range(c(ref[,2], target[,2])) * 1.1
  
  # Set up the plot
  plot(ref[,1], ref[,2],
       asp = 1,
       pch = 16,
       col = "pink",
       xlim = xlim_range,
       ylim = ylim_range,
       xlab = "X",
       ylab = "Y",
       main = paste("Ref:", pair[1], "→ Target:", pair[2]))
  
  # Add the target points
  points(target[,1], target[,2], 
         pch = 17, 
         col = "purple")
  
  # Add a cleaner legend
  legend("bottomleft", 
         legend = c(paste("Ref:"), paste("Target:")), 
         col = c("pink", "purple"), 
         pch = c(16, 17),
         pt.cex = 1.5,    # bigger points in legend
         cex = 0.8,       # shrink text slightly
         bty = "n",       # no box around the legend
         inset = 0.02)    # slight inset to avoid the edges
}

```